
---
title: "Q3"
output: html_document
---

## 1.
in this analysis, we will examine the **quarterly U.S. Gross National Product (GNP)** data using two time series models:  
1. **AR(1)** (autoregressive of order 1)  
2. **ARMA(1,2)** (autoregressive-moving average model of order 1 and 2)

i explore these models on the **differenced logarithm of the GNP** data.  
goal is to:

1. Perform detailed **model diagnostics** for both models.

2. **Compare** the two models based on diagnostic results, using AIC values, residual checks, and plots.

---

## 2.AR(1) and ARMA(1,2)

### 2.1 AR(1) Model

The AR(1) model is defined as:
\[
X_t = \phi X_{t-1} + W_t,
\]
where:
- \(X_t\) is the current value of the time series.
- \(\phi\) is the AR(1) coefficient (captures the dependence on the previous value).
- \(W_t\) is white noise with zero mean and constant variance \(\sigma_W^2\).

The AR(1) model assumes that each observation is linearly related to the previous one, making it suitable for **persistent time series with slow decay** in autocorrelations.

### 2.2 ARMA(1,2) Model

The ARMA(1,2) model is formulated as:
\[
X_t = \phi X_{t-1} + W_t + 	\theta_1 W_{t-1} + 	\theta_2 W_{t-2},
\]
where:
- \(W_t, W_{t-1}, W_{t-2}\) are white noise terms.
- \(	\theta_1, 	\theta_2\) are MA coefficients capturing the short-term effects of noise on the series.

This model accounts for both **long-term dependencies** (through AR terms) and **short-term shocks** (through MA terms).

---

## 3. Model Diagnostics: Key Steps

For both AR(1) and ARMA(1,2) models, we perform:
1. **Parameter estimation**: Estimate AR and MA coefficients.
2. **Residual analysis**: Check if residuals behave like white noise.
3. **Autocorrelation checks**: Use ACF and PACF plots to validate the model.
4. **Model selection**: Compare models using **AIC** (Akaike Information Criterion).

---

## 4. Load Required Libraries and Data

```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
library(forecast)
library(tseries)
library(astsa)

# Load and preprocess the data
data("gnp")
gnp_diff <- diff(log(gnp))  # Differenced log GNP data
```

---

## 5. Fitting the AR(1) Model

We now fit an **AR(1)** model to the differenced log GNP data.

```{r}
# Fit AR(1) model
ar1_model <- arima(gnp_diff, order = c(1, 0, 0))

# Summary of the AR(1) model
summary(ar1_model)

# Diagnostics plots for AR(1) model
tsdiag(ar1_model)

# Check residuals for normality and autocorrelation
checkresiduals(ar1_model)
```

### 5.1 Interpretation of AR(1) Model Results

1. **Estimated Parameters**: The AR(1) coefficient \( \phi \) captures the relationship with the previous time step.
2. **Residual Analysis**: 
   - **Ljung-Box Test**: If the p-value is greater than 0.05, residuals are uncorrelated.
   - **ACF/PACF Plots**: These plots help confirm if there is any remaining autocorrelation.
   - **Normality**: Check if residuals are normally distributed using Q-Q plots and histograms.

---

## 6. Fitting the ARMA(1,2) Model

We now fit an **ARMA(1,2)** model to the same data.

```{r}
# Fit ARMA(1,2) model
arma12_model <- arima(gnp_diff, order = c(1, 0, 2))

# Summary of the ARMA(1,2) model
summary(arma12_model)

# Diagnostics plots for ARMA(1,2) model
tsdiag(arma12_model)

# Check residuals for normality and autocorrelation
checkresiduals(arma12_model)
```

### 6.1 Interpretation of ARMA(1,2) Model Results

1. **Estimated Parameters**: Includes both AR(1) and MA(2) coefficients.
2. **Residual Analysis**:
   - **ACF/PACF**: Check if residuals are white noise.
   - **Ljung-Box Test**: Used to verify if residuals are uncorrelated.
   - **Normality Check**: Evaluate residuals for normality.

---

## 7. Comparison of AR(1) and ARMA(1,2) Models

```{r}
# Compare AIC values for both models
aic_ar1 <- AIC(ar1_model)
aic_arma12 <- AIC(arma12_model)

cat("AIC for AR(1):", aic_ar1, "
")
cat("AIC for ARMA(1,2):", aic_arma12, "
")
```

### 7.1 Model Comparison Summary

1. **AIC Comparison**: 
   - The model with the lower AIC is preferred as it provides a better balance between model fit and complexity.
   - If AR(1) has a lower AIC, it indicates that a simpler model is sufficient.

2. **Residual Diagnostics**: Both models should have uncorrelated residuals with no significant autocorrelations.

3. **Interpretability**: AR(1) is simpler and easier to interpret compared to the more complex ARMA(1,2) model.

---

## 8. Conclusion

In this analysis, both AR(1) and ARMA(1,2) models fit the differenced log GNP data reasonably well.  
- **AR(1)** model offers a simpler interpretation and may be preferred if AIC values are similar.
- **ARMA(1,2)** captures more complex relationships but introduces additional parameters.

Based on the results, we recommend the AR(1) model for its simplicity unless the ARMA(1,2) model shows a significantly better fit.
