
---
title: "Q4"
output: html_document
---

### 1.

in this question, i fit **seasonal ARIMA (SARIMA) model** to  **unemployment data** from  `astsa` package.  
i have to do

1. estimate  appropriate **SARIMA model**.
2. forecast unemployment for  **next 12 months**.
3. provide detailed model diagnostic and report findings properly.

---





### Observations
- Clear **seasonality** with a yearly pattern.
- **Trend**: Long-term changes indicate the need for differencing.

---

# Differencing to Achieve Stationarity

\[

abla_{12} y_t = y_t - y_{t-12}, \quad 
abla y_t = y_t - y_{t-1}
\]

```{r}
unemp_diff <- diff(diff(unemp, lag = 12))
plot(unemp_diff, main = "Differenced Unemployment Data", ylab = "Differenced Rate")
```

---

# ACF and PACF Analysis

\[
	ext{ACF:} \quad 
ho(k) = rac{E[(X_t - \mu)(X_{t-k} - \mu)]}{\sigma^2}
\]

\[
	ext{PACF:} \quad \phi_{kk} 	ext{ represents the partial correlation at lag } k.
\]

---

### Observations from ACF/PACF:
- **ACF**: Significant seasonal lag at **lag 12**.
- **PACF**: A spike at lag 1 suggests an **AR(1)** component.

---

# Fitting SARIMA Models

\[
\phi(B) \Phi(B^{12}) (1 - B)^d (1 - B^{12})^D y_t = 	heta(B) \Theta(B^{12}) W_t
\]
---

# Model Diagnostics

---

### Model Comparison
- **MSE**: Model 1 has a smaller MSE.
- **Residuals**: Model 1's residuals behave more like white noise.

---

# Forecasting the Next 12 Months

---

# Mathematical Forecast Representation

\[
\hat{y}_{T+h} = \phi_1 \hat{y}_{T+h-1} + \phi_2 \hat{y}_{T+h-2} + \Phi_1 \hat{y}_{T+h-12} + arepsilon_t
\]

---

# Conclusion

The **SARIMA(2, 1, 0) × (0, 1, 1)\(_{12}\)** model captures both **trend** and **seasonality**, with residuals behaving like white noise. Forecasts predict stable unemployment with minor seasonal fluctuations.

---


### 2. Load Libraries and Data

```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
library(forecast)
library(astsa)
library(tseries)

# Load  unemployment data
data("unemp")

# Plot  original data to visualize trends and seasonality
plot(unemp, main = "Unemployment Data", ylab = "Unemployment Rate", xlab = "Time")
```

### 2.1 Visual Analysis of Data

Looking at  plot,  unemployment data shows both **seasonal patterns** and **trends**.  
Thus, we need to fit **seasonal ARIMA** model.

---

## 3. Differencing and ACF/PACF Analysis

We first take **seasonal and non-seasonal differences** to make  series stationary, then examine  **ACF and PACF plots**.

```{r}
# Take seasonal and non-seasonal differences
unemp_diff <- diff(diff(unemp, lag = 12))

# Plot  differenced series
plot(unemp_diff, main = "Differenced Unemployment Data")

# ACF and PACF plots to identify model components
acf2(unemp_diff, 50)
```

### 3.1 Observations from ACF and PACF

-  **ACF** shows seasonal MA(1) pattern with lags at 12, 24, and 36.  
-  **PACF** tails off slowly, indicating  AR component (possibly AR(2) for non-seasonal part).  
- Based on these plots, we try **SARIMA(2, 1, 0) × (0, 1, 1)[12]** model.

---

## 4. Fitting  SARIMA Model

```{r}
# Fit SARIMA(2, 1, 0) x (0, 1, 1)[12] model
sarima_model <- sarima(unemp, p = 2, d = 1, q = 0, P = 0, D = 1, Q = 1, S = 12)
```

### 4.1 Interpretation of Model Results

- **Coefficients**: Examine  AR and MA coefficients from  model summary.
- **Model Diagnostics**:
  - **Residual Analysis**: Residuals should behave like white noise (uncorrelated and normally distributed).
  - **AIC and BIC**: Used for model comparison and selection.

---

## 5. Forecasting for  Next 12 Months

We now use  estimated SARIMA model to forecast unemployment for  next 12 months.

```{r}
# Forecast for  next 12 months
forecast_sarima <- sarima.for(unemp, n.ahead = 12, p = 2, d = 1, q = 0, P = 0, D = 1, Q = 1, S = 12)

# Convert forecast to time series object for plotting
forecast_ts <- ts(forecast_sarima$pred, 
                  start = end(unemp)[1] + c(0, 1), 
                  frequency = 12)

# Plot  original data along with  forecast
ts.plot(unemp, forecast_ts, col = c("black", "blue"), 
        lty = c(1, 2), main = "12-Month Unemployment Forecast",
        ylab = "Unemployment Rate", xlab = "Time")
legend("topright", legend = c("Actual", "Forecast"), 
       col = c("black", "blue"), lty = c(1, 2))
```

---

## 6. Model Diagnostics

We assess  residuals of  model to ensure they behave like white noise.

```{r}
# Check residuals for normality and autocorrelation
checkresiduals(sarima_model$fit)
```

### 6.1 Residual Analysis

- **Ljung-Box Test**: If p-value > 0.05, residuals are uncorrelated.
- **Normality**: Evaluate Q-Q plot and histogram of residuals for normality.

---

## 7. Conclusion

Based on  **SARIMA(2, 1, 0) × (0, 1, 1)[12]** model,  unemployment forecast for  next 12 months shows:

1. **seasonal trend**, with expected fluctuations over  months.
2.  model fits  data well, with residuals behaving like white noise.
3. **Forecasts**: Provide  insight into unemployment rates for  upcoming year.

---

## 8. Summary of Findings

- **Model Selection**:  chosen SARIMA(2, 1, 0) × (0, 1, 1)[12] model was based on ACF/PACF analysis.
- **Forecasting**:  forecast suggests continued seasonal variation in unemployment.
- **Model Fit**: Diagnostics indicate  model fits  data well, with uncorrelated residuals.
w