
---
title: "Q5"
output: html_document
---

## 1. Introduction

In this task, we analyze the **Johnson & Johnson (J&J) quarterly earnings** dataset using a **Seasonal ARIMA (SARIMA) model**.  
The goal is to:
1. **Log-transform** the data to stabilize the variance.
2. Apply **seasonal differencing** to make the data stationary.
3. Fit an appropriate **SARIMA model** to the data.
4. **Forecast the next 4 quarters** and evaluate the model's performance.

---

## 2. Load Libraries and Data

```{r, warning=FALSE, message=FALSE}
# Load required libraries
library(astsa)
library(forecast)

# Load the Johnson & Johnson earnings data
data("jj")

# Plot the original data
plot(jj, type = "o", main = "Johnson & Johnson Quarterly Earnings",
     ylab = "Quarterly Earnings per Share", xlab = "Time")
```

### 2.1 Visual Analysis of Data

The plot of the original data shows both **trend** and **seasonal patterns**, with increasing variability over time.  
Thus, it is appropriate to **log-transform** the data to stabilize the variance.

---

## 3. Log Transformation and Differencing

We take the **log of the data** to stabilize the variance and apply **first and seasonal differencing** to make it stationary.

```{r}
# Log-transform the data
log_jj <- log(jj)

# Apply first and seasonal differencing
diff_log_jj <- diff(diff(log_jj, lag = 4))

# Plot the differenced series
plot(diff_log_jj, main = "Differenced Log-transformed J&J Data")
acf2(diff_log_jj, 50)  # ACF and PACF plots
```

### 3.1 Observations from ACF and PACF

- **ACF**: Seasonal lags at 4, 8, 12, indicating a **seasonal component**.
- **PACF**: Suggests an AR(1) component with some seasonal correlation.
- We choose to fit a **SARIMA(1,1,0) × (1,1,0)[4]** model based on these observations.

---

## 4. Fitting the SARIMA Model

```{r}
# Fit the SARIMA(1,1,0) × (1,1,0)[4] model
sarima_model <- sarima(log_jj, p = 1, d = 1, q = 0, P = 1, D = 1, Q = 0, S = 4)
```

### 4.1 Model Diagnostics

- **Coefficients**: Review the AR and MA coefficients from the model summary.
- **Residual Analysis**: Residuals should be white noise.
- **AIC/BIC**: Used for model comparison.

---

## 5. Forecasting the Next 4 Quarters

We now forecast the **next 4 quarters** using the fitted SARIMA model.

```{r}
# Forecast the next 4 quarters
forecast_sarima <- sarima.for(log_jj, n.ahead = 4, p = 1, d = 1, q = 0, P = 1, D = 1, Q = 0, S = 4)

# Convert forecast to time series object
forecast_values <- ts(forecast_sarima$pred, start = end(jj)[1] + c(0, 1), frequency = 4)

# Plot original data with forecast
ts.plot(jj, exp(forecast_values), col = c("black", "red"), lty = c(1, 2),
        main = "4-Quarter Forecast of J&J Earnings", ylab = "Earnings", xlab = "Time")
legend("topright", legend = c("Actual", "Forecast"), col = c("black", "red"), lty = c(1, 2))
```

---

## 6. Model Diagnostics

We assess the residuals to ensure the model fits well.

```{r}
# Check residuals for normality and autocorrelation
checkresiduals(sarima_model$fit)
```

### 6.1 Residual Analysis

- **Ljung-Box Test**: Residuals should be uncorrelated (p-value > 0.05).
- **Q-Q Plot**: Check if residuals are normally distributed.

---

## 7. Conclusion

Based on the **SARIMA(1,1,0) × (1,1,0)[4]** model, the forecast for the next 4 quarters suggests:

1. A continuation of the seasonal pattern in earnings.
2. The model fits well, with residuals behaving like white noise.
3. **Forecasts** provide insights into future earnings trends.

---

## 8. Summary of Findings

- **Model Selection**: The SARIMA(1,1,0) × (1,1,0)[4] model was chosen based on ACF/PACF analysis.
- **Forecasting**: The forecast suggests continued seasonal variations in earnings.
- **Model Fit**: Diagnostics indicate the model fits well, with uncorrelated residuals.
